---
title: "Question 3"
author: "Maia and Yudong"
date: "2025-11-25"
output: pdf_document
---


```{r}
# Load required libraries
library(tidyverse)
library(MASS)
library(caret)
library(ggplot2)
library(randomForest)
library(pROC)
library(dplyr)
```

```{r}
# Load the scaled data
dat<- read.csv("scaleddata.csv")
head(dat)
summary(dat)
```

```{r}
names(dat)

#check what infection classes are present
unique(dat$Class.name)

#check missing values per column
colSums(is.na(dat))

```

```{r}
# PCA Analysis to Explore Demographic Influences
set.seed(111)
cat("Scaled data:", dim(dat), "\n")
dat$Age <- gsub("[^0-9.]", "", dat$Age)
dat$Age <- as.numeric(dat$Age)
mean_age <- mean(dat$Age, na.rm = TRUE)
dat$Age[is.na(dat$Age)] <- mean_age
scaled_clean <- dat %>% filter(!is.na(Sex))
scaled_clean$Sex <- as.factor(scaled_clean$Sex)
metabolites <- scaled_clean[, 7:ncol(scaled_clean)]
```

```{r}
# Perform PCA and plot for visualization
pca <- prcomp(metabolites, scale. = TRUE)
pca_df <- data.frame(pca$x[, 1:5],
                     Sex = scaled_clean$Sex,
                     Age = scaled_clean$Age)
#Create Age groups for shape
pca_df$AgeGroup <- cut(
  pca_df$Age,
  breaks = c(-Inf, 30, 60, Inf),
  labels = c("Young", "Middle", "Older")
)

# PCA visualization 
ggplot(pca_df, aes(x = PC1, y = PC2, 
                   color = Sex, 
                   shape = AgeGroup)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal(base_size = 14) +
  labs(title = "PCA of Scaled Metabolomic Data by Sex and Age",
       subtitle = "Using standardized metabolite abundance only",
       x = "PC1", y = "PC2",
       shape = "Age Group") +
  scale_color_manual(values = c("Female" = "red", "Male" = "skyblue")) +
  scale_shape_manual(values = c(16, 17, 15))

```
Figure 4. Linear regression results showing that age explains very little variation in the first two principal components (PC1 and PC2). PC1 shows a weak but statistically significant association with age (R² ≈ 0.02, p ≈ 0.039), while PC2 shows no detectable relationship (R² < 0.001, p ≈ 0.81). These findings indicate that age does not strongly contribute to the major axes of metabolomic variation.

```{r}
# Linear regression of PC1 and PC2 against Age
lm_PC1 <- lm(PC1 ~ Age, data = pca_df)
lm_PC2 <- lm(PC2 ~ Age, data = pca_df)
summary(lm_PC1)
summary(lm_PC2)
```

```{r}
# Scatter plot of Age vs PC1 with regression line
ggplot(pca_df, aes(Age, PC1)) +
  geom_point(alpha = 0.7, color = "skyblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  theme_minimal(base_size = 14) +
  labs(title = "Relationship Between Age and PC1",
       subtitle = "Expected weak linear correlation (Scaled data only)",
       x = "Age", y = "PC1 score")
```
Figure 5. This scatter plot illustrates the association between participant age and the first principal component (PC1) derived from scaled metabolomic features. Each point represents an individual sample, and the fitted linear regression line (red) shows a weak positive trend (β ≈ small, p ≫ 0.05), indicating that age explains very little of the variation in PC1. This suggests that age does not substantially contribute to the major axis of metabolomic variation in this dataset.

```{r}
#Check Sex and Age
table(pca_df$Sex, useNA = "ifany")
summary(pca_df$Age)
```
```{r}

#LDA For Sex
lda_sex<- lda(Sex~PC1+PC2+PC3+PC4+PC5,
              data = pca_df,
              CV= TRUE)
sex_pred<- lda_sex$class
sex_acc<- mean(sex_pred==pca_df$Sex)
sex_acc
table(True=pca_df$Sex, Pred= sex_pred)

#Fit LDA model to get LD scores 
set.seed(111)
lda_sex_fit<- lda(Sex~PC1+PC2+PC3+PC4+PC5, data = pca_df)
lda_sex_scores<- predict(lda_sex_fit)$x
lda_sex_df<- data.frame(lda_sex_scores, Sex=pca_df$Sex)

```

```{r}
#Plot LD1 vs LD2 by Sex
ggplot(lda_sex_df, aes(x=LD1, fill = Sex)) +
  geom_density(alpha=0.5) +
  theme_minimal(base_size = 14) +
  labs(title = "LDA of Metabolomic PCs by Sex", 
       subtitle = "Density of LD1 scores by sex", x = "LD1", y = "Density")
```
Figure 6. Linear Discriminant Analysis (LDA) of the first five principal components reveals sex-associated structure in nasal metabolite profiles. The density plot shows the distribution of LD1 scores for female (pink) and male (blue) samples; males are shifted toward higher LD1 values, with partial overlap between sexes, indicating some separation of metabolomic profiles by sex.
```{r}
#LDA for Age

median_age<- median(pca_df$Age, na.rm=TRUE)
pca_df$AgeGroup<-ifelse(pca_df$Age <=median_age, "Younger", "Older")
pca_df$AgeGroup<-factor(pca_df$AgeGroup)
table(pca_df$AgeGroup)

lda_age_fit<- lda(Age~PC1+PC2+PC3+PC4+PC5, data = pca_df)
lda_age_scores<- predict(lda_age_fit)$x
lda_age_df<- data.frame(lda_age_scores, AgeGroup=pca_df$AgeGroup)

ggplot(lda_age_df, aes(x=LD1, fill= AgeGroup)) +
  geom_density(alpha=0.5) +
  theme_minimal(base_size = 14) +
  labs(title = "LDA of Metabolomic PCs by Age", 
       subtitle = "Density of LD1 scores by age", x= "LD1", y= "Density")

```
Figure 7. Linear Discriminant Analysis (LDA) of metabolomic principal components by age group. Density plots of LD1 scores for younger (<median age (27 years old) in blue) and older (> median (27 years old, pink) participants show highly overlapping distributions, indicating little evidence that overall metabolomic profiles differ with age. 

